---
title: "InDen T cells | Creation of HL vgm object and gene pairing computation"
author: "Giorgio Gonnella"
date: "2024-06-28"
output:
  html_document:
    theme: sandstone
params:
  prjpath: "/srv/baia/inden/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
  mixcr: "/home/giorgio/software/mambaforge/pkgs/mixcr-3.0.12-0/share/mixcr-3.0.12-0/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=T)
.libPaths(c(params$libpath, .libPaths()))

library(Platypus)
library(dplyr)

analysis_root <- paste0(params$prjpath, "analysis-1/")

encode_input <- paste0(params$prjpath, "data/bcr_vdj_output")

input_step_dir <- paste0(analysis_root, "steps/110.0.P.B.BCR_analysis.platypus/")
input_var_dir <- paste0(input_step_dir, "results/vars/")
if (!dir.exists(input_var_dir)) {
  stop("Input variable directory does not exist: ", input_var_dir)
}

this_step_dir <- paste0(analysis_root, "steps/110.0.P.B.BCR_analysis.platypus/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}

var_dir <- paste0(this_step_dir, "results/vars/")
dir.create(var_dir, showWarnings = FALSE, recursive = TRUE)

plots_dir <- paste0(this_step_dir, "results/plots/")
dir.create(plots_dir, showWarnings = FALSE, recursive = TRUE)

helpers_dir <- paste0(params$prjpath, "helpers/")
source(paste0(helpers_dir, "platypus_helpers.R"))
```

# Load data

```{r load_vgm}
vgm_HL <- readRDS(file = paste0(input_var_dir, "vgm.rds"))
```

## Filter cells wo VJ

We filter here the cells without VJ and save them as a separate object vgm\_HL.

```{r filter_unavailable_vj}
vgm_HL <- vgm_filter_unavailable_vj(vgm_HL)
```

```{r cmp_n_cells_after_filter_unavail_VJ}
gex_data <- data.frame(sample = factor(vgm_HL[["GEX"]]$sample), library = "GEX",
                       celltype = factor(vgm_HL[["GEX"]]$celltype))
vdj_data <- data.frame(sample = factor(vgm_HL[["VDJ"]]$sample), library = "VDJ",
                       celltype = factor(vgm_HL[["VDJ"]]$celltype))
combined_data <- bind_rows(gex_data, vdj_data)
combined_plot <- ggplot(combined_data,
  aes(x = library, y = ..count.., fill=celltype)) +
  geom_bar(stat = "count") +
  facet_wrap(~sample,nrow=1)+
  xlab("Sample") + ylab("N.cells") +
  ggtitle("Number of cells") +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = vgm_HL[["GEX"]]@misc$fctype_mod_cols)
print(combined_plot)
ggsave(paste0(plots_dir, "after_filter_unavail_VJ.n_cells_per_sample.pdf"),
       combined_plot)
```

### Table GEX/VDJ available after filtering unavailable VJ chains

```{r n_unavailable_after_filter_unavailable_vj}
print("VDJ available for cells in GEX dataset:")
table(vgm_HL[["GEX"]]$VDJ_available)

print("VDJ available for cells in GEX dataset, by sample:")
table(vgm_HL[["GEX"]]$sample, vgm_HL[["GEX"]]$VDJ_available)

print("VDJ available for cells in GEX dataset, by celltype:")
vgm_HL[["GEX"]]$celltype <- droplevels(vgm_HL[["GEX"]]$celltype)
vgm_HL[["GEX"]]$celltype.abbrev <- droplevels(vgm_HL[["GEX"]]$celltype.abbrev)
table(vgm_HL[["GEX"]]$celltype, vgm_HL[["GEX"]]$VDJ_available)

print("VDJ available for cells in GEX dataset, by cluster:")
table(vgm_HL[["GEX"]]$seurat_clusters, vgm_HL[["GEX"]]$VDJ_available)
```

## Clonotypes

Recompute for each cell the size of the clonotype to which it belongs.
Group into bins for plotting.

```{r clonotype_size_vgm_HL}
vgm_HL <- cells_clonotype_size(vgm_HL)
vgm_HL <- compute_clonotype_size_bins(vgm_HL)
```

## Gene usage analysis

```{r gene_usage_analysis_all_data}
all_data_results <- gene_usage_signif(vgm_HL$VDJ, genesets = c("VH_VL_pairing"))
```

```{r gene_usage_analysis_cell_types, message=T}
cell_types <- unique(vgm_HL$VDJ$pruned.labels_singleR_fine.mod)
cell_type_results <- data.frame()
for (ct in cell_types) {
  message("Computing gene usage for cell type: ", ct)
  results <-
    gene_usage_signif(vgm_HL$VDJ, celltype = ct,
                      celltype_column = "pruned.labels_singleR_fine.mod",
                      genesets = c("VH_VL_pairing"),
                      n_permutations = 10000)
  if (!is.null(results) && nrow(results) > 0) {
    cell_type_results <- rbind(cell_type_results, results)
  }
}
```

```{r gene_usage_analysis_unsplitted}
unsplitted_results <-
  gene_usage_signif(vgm_HL$VDJ,
                    celltype = "Plasmablasts",
                    celltype_column = "pruned.labels_singleR_fine",
                    genesets = c("VH_VL_pairing"))
```

```{r gene_usage_analysis_merge}
vgm_HL$GEX@misc$gene_pairings$all <- rbind(all_data_results,
                                           cell_type_results,
                                           unsplitted_results)
vgm_HL$GEX@misc$gene_pairings$p_filt <-
  vgm_HL$GEX@misc$gene_pairings$all[
    vgm_HL$GEX@misc$gene_pairings$all$p_adj < 0.05,]
```

#### Gene usage analysis hospitalized plasmablasts

```{r}
plasmablasts <- subset(vgm_HL$VDJ,
                       celltype %in% c("Proliferating plasmablasts",
                                       "Non-proliferating plasmablasts"))
plasmablasts$celltype = droplevels(plasmablasts$celltype)
results <- gene_usage_signif(plasmablasts, celltype = "Hospitalized",
                             celltype_column = "clinical_condition",
                             group_column = "celltype",
                             genesets = c("VH_VL_pairing"),
                             n_permutations=10000)
signif_results <- results[results$p_adj < 0.05, ]

vgm_HL$GEX@misc$gene_pairings$hpb_all <- results
vgm_HL$GEX@misc$gene_pairings$hpb_p_filt <- signif_results
```

## SHM analysis

```{r}
vgm_HL$VDJ$VDJ_SHM <- NULL
vgm_HL$VDJ$VJ_SHM <- NULL
files <- list.files(pattern = "^temp")
file.remove(files)
mixcr_out <- VDJ_call_MIXCR(
    VDJ = vgm_HL$VDJ,
    mixcr.directory = params$mixcr,
    species = "hsa",  # or "mmu" for mouse
    platypus.version = "v3",
    operating.system = "Linux",
    simplify = TRUE,
    custom.cmd.call = "java -jar"
)
vgm_HL$VDJ <- mixcr_out
```

## Save

```{r saving_vgm_HL}
saveRDS(vgm_HL, file = paste0(var_dir, "vgm_HL.rds"))
```

## Session Info

```{r session_info, echo=FALSE}
sessionInfo()
```

