---
title: "InDen B cells | Integration using Harmony method"
author: "Giorgio Gonnella"
date: 2024-06-27
output:
  html_document:
    theme: sandstone
params:
  prjpath: "/srv/baia/inden/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
  input_merged_rdata: FALSE
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=F)
.libPaths(c(params$libpath, .libPaths()))

library(harmony)
library(Seurat)

analysis_root <- paste0(params$prjpath, "analysis-1/")

if (params$input_merged_rdata) {
  input_var_dir <- paste0(params$prjpath, "merged/B/")
} else {
  prev_step_dir <- paste0(analysis_root, "steps/103.0.P.B.merging.seurat/")
  input_var_dir <- paste0(prev_step_dir, "results/vars/")
}
if (!dir.exists(input_var_dir)) {
  stop("Input variable directory does not exist: ", input_var_dir)
}

this_step_dir <- paste0(analysis_root, "steps/104.0.P.B.integration.harmony/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}

var_dir <- paste0(this_step_dir, "results/vars/")
dir.create(var_dir, showWarnings = FALSE, recursive = TRUE)

helpers_dir <- paste0(params$prjpath, "helpers/")
source(paste0(helpers_dir, "clustering_helpers.R"))
```

```{r load_file, echo = FALSE}
if (params$input_merged_rdata) {
  load(paste0(input_var_dir, "TC_SC03_allbcells.Rdata"))
  so <- TC_SC03_allbcells
} else {
  so <- readRDS(paste0(input_var_dir, "merged_so.rds"))
}
```

In the integration step, the Harmony algorithm is used to integrate the data
from the different samples. It is a method which aims to correct for batch
effects while preserving the biological signal.

After the integration, the data is clustered and visualized using UMAP.

# Running Harmony

```{r run_harmony, echo = TRUE}
set.seed(154)
so <- RunHarmony(object = so,
                 group.by.vars=c("orig.ident"),
                 reduction = "pca", reduction.save="harmony")
```

# Repeat analysis pipeline after integration

## UMAP

```{r setting_umap_parameters, echo=TRUE}
myDims <- 1:40
```

```{r umap , message=FALSE, warning=FALSE}
set.seed(154)
so <- RunUMAP(object = so, dims = myDims,
              reduction = "harmony", reduction.name = "umapharm")
```

## Clustering

```{r setting_clustering_parameters, echo=TRUE}
reS <- c(0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8)
```

```{r clustering, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(154)
so <- FindNeighbors(object = so, dims = myDims,
                        reduction = "harmony", graph.name = "harmsnn")
so <- FindClusters(object = so, resolution = reS,
                        graph.name = "harmsnn")
so <- factor_resolution(so)
```

# Saving data

```{r save, echo=TRUE}
saveRDS(so, file = paste0(var_dir, "integrated_so.raw.rds"))
```

# Session info

```{r session_info}
sessionInfo()
```
