---
title: "InDen B cells | Merging individual samples to a single Seurat object"
author: "Giorgio Gonnella"
date: 2024-04-29
output: html_document
params:
  pfx: "/srv/baia/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
  input_prefilt_rdata: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=F)
.libPaths(c(params$libpath, .libPaths()))

library(Seurat)

prj_root <- paste0(params$pfx, "inden/")
analysis_root <- paste0(prj_root, "analysis-1/")

prev_step_dir <- paste0(analysis_root, "steps/102.0.P.B.qc_filter/")
input_var_dir <- paste0(prev_step_dir, "results/vars/")
if (!dir.exists(input_var_dir)) {
  stop("Input variable directory does not exist: ", input_var_dir)
}

this_step_dir <- paste0(analysis_root, "steps/103.0.P.B.merging.seurat/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}
var_dir <- paste0(this_step_dir, "results/vars/")
dir.create(var_dir, showWarnings = FALSE, recursive = TRUE)

helpers_dir <- paste0(prj_root, "helpers/")
source(paste0(helpers_dir, "merging_helpers.R"))
source(paste0(helpers_dir, "clustering_helpers.R"))
```

In the merging step, the individual samples are merged into a single
Seurat object. The UMAP, PCA, and clustering are performed on the merged
object.

## Importing the Seurat objects

The files are imported as a list of Seurat objects.

```{r importing, echo=TRUE}
if (params$input_prefilt_rdata) {
  stop("Re-implement this copying from version tagged v1")
}
bcells_files <- list.files(path = input_var_dir,
                           pattern = "_so_filtered.rds")
bcells_ls <- sapply(bcells_files, function(x)
                    readRDS(paste0(input_var_dir, x)))
names(bcells_ls) <- gsub("_so_filtered.rds", "", bcells_files)
rm(bcells_files)
```

## Filtering

Sample 10 is removed because of the low number of cells.

```{r datasets_filtering}
exclude <- c("sample_10_B")
keep <- setdiff(names(bcells_ls), exclude)
bcells_ls <- bcells_ls[keep]
```

## Merging

```{r prepare_merging}
bcells_ls <- lapply(bcells_ls, function (so) {
  orig.sample_id <- sub(".*S(\\d+)B.*", "\\1", so@project.name)
  colnames(so) <- paste0(colnames(so), "_", orig.sample_id)
  DefaultAssay(so) <- "RNA"
  return(so)
})
```

The Seurat objects for the individual samples are merged
into one Seurat object, using the `merge()` function.

```{r merging, echo=TRUE}
so <- merge(x = bcells_ls[[1]], y = bcells_ls[-1], project = "bcells")
if (so_version(so) == "5") {
  so <- JoinLayers(so)
}

rm(list = (c("bcells_ls")))
```

The samples in the merged object are ordered by sample number.

```{r sample_ordering, echo=TRUE}
so$orig.ident[
      so$orig.ident == "sample_15_B_so"] <- "TC_SC03_S15B_so"
so$orig.ident[
      so$orig.ident == "sample_16_B_so"] <- "TC_SC03_S16B_so"
so$orig.ident <- factor(so$orig.ident, levels = c("TC_SC03_S1B_so",
                                                  "TC_SC03_S5B_so",
                                                  "TC_SC03_S6B_so",
                                                  "TC_SC03_S7B_so",
                                                  "TC_SC03_S8B_so",
                                                  "TC_SC03_S9B_so",
                                                  "TC_SC03_S11B_so",
                                                  "TC_SC03_S12B_so",
                                                  "TC_SC03_S13B_so",
                                                  "TC_SC03_S15B_so",
                                                  "TC_SC03_S16B_so"))
so$sample <- sub(".*(S\\d+B).*", "\\1", so$orig.ident)
```

Add clinical condition metadata.

```{r add_clinical_condition, echo=TRUE}
conditions <- c("Hospitalized", "Subclinical")
hosp_n = c("1", "2", "3", "5", "6", "7", "15", "16")
so$clinical_condition <-
  ifelse(so$orig.ident %in% paste0("TC_SC03_S", hosp_n, "B_so"),
         conditions[1], conditions[2])
```

## Normalization

```{r normalization, echo=TRUE}
so <- NormalizeData(object = so, assay = "RNA")
```

## Selection of highly variable genes

```{r hvg_selection, echo=TRUE}
so <- FindVariableFeatures(object = so, assay = "RNA",
                           nfeatures = 2500, selection.method = "vst")
```

Removing unwanted source of variation in the highly variable genes

```{r correcting_hvg, echo=TRUE}
if (params$input_prefilt_rdata) {
  vf <- so@assays$RNA@var.features
} else {
  vf <- VariableFeatures(so)
}

unwanted_tcr <- grep(pattern = "^TR[AB][VC]", vf, value = TRUE)
unwanted_ig <- c(grep(pattern = "^IG[HLK][VDJ]", vf, value = TRUE),
              c("IGHM", "IGHD", "IGHE", "IGHA[12]", "IGHG[1234]", "IGKC", "IGLC[1234567]", "AC233755.1"))
unwanted_mt <- grep(pattern = "^MT", vf, value = TRUE)
unwanted_ribo <- grep(pattern = "^RP[LS]", vf, value = TRUE)
unwanted_vf <- c(unwanted_tcr, unwanted_ig, unwanted_mt, unwanted_ribo)
hvg_corr <- vf[!vf %in% unwanted_vf]
```

## Scaling data

```{r scaling, echo=TRUE}
so <- ScaleData(object = so,
            features = vf,
            vars.to.regress = c("nCount_RNA", "percent.mito"),
            assay = "RNA")
```


## Linear dimension reduction (PCA)

```{r pca, echo=TRUE}
so <- RunPCA(object = so, features = hvg_corr,
             npcs = 50, assay = "RNA")
```

## Non-linear dim reduction (UMAP)

```{r umap, echo=TRUE}
myDims <- 1:40
so <- RunUMAP(object = so, dims = myDims,
              reduction = "pca", reduction.name = "umap")
```

## Clustering

```{r clustering, echo=TRUE}
reS <- c(0.2, 0.4, 0.6, 0.8, 1.0, 1.2)
so <- FindNeighbors(object = so, dims = myDims)
so <- FindClusters(object = so, resolution = reS)
so <- factor_resolution(so)
```

## Saving Data

```{r saving}
saveRDS(so, paste0(var_dir, "merged_so.rds"))
```

## Session Info

```{r session_info, echo=FALSE}
sessionInfo()
```
