---
title: "InDen T cells | Split Effector Memory CD8+ cells"
author: "Giorgio Gonnella"
date: 2024-05-21
output:
  html_document:
    clean: FALSE
    theme: sandstone
params:
  pfx: "/srv/baia/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=T)
.libPaths(c(params$libpath, .libPaths()))

library(Seurat)
library(scProportionTest)

prj_root <- paste0(params$pfx, "inden/")
analysis_root <- paste0(prj_root, "analysis-1/")

prev_step_dir <- paste0(analysis_root, "steps/004.0.P.T.integration.harmony/")
input_var_dir <- paste0(prev_step_dir, "results/vars/")
if (!dir.exists(input_var_dir)) {
  stop("Input variable directory does not exist: ", input_var_dir)
}

this_step_dir <- paste0(analysis_root, "steps/008.0.P.T.split_CD8EM/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}

var_dir <- paste0(this_step_dir, "results/vars/")
dir.create(var_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r load_data, echo = FALSE}
so <- readRDS(paste0(input_var_dir, "integrated_so.rds"))
```

The purpose of this step is to divide the effector memory CD8 cells according
to their unsupervised clustering. The cluster 11 cells will be separated from
the other cells, and called, respectively,
"Proliferating effector memory CD8 T cells" and
"Non-proliferating effector memory CD8 T cells".

# Before splitting

```{r before, echo=F}
table(so$pruned.labels_singleR_fine)
```

# Splitting

```{r splitting, echo = TRUE}
g.b = so@misc$sel_res_col
sel_cluster = "11"

so$pruned.labels_singleR_fine.mod <-
  so$pruned.labels_singleR_fine
so$pruned.labels_singleR_fine.mod <-
  ifelse(so$pruned.labels_singleR_fine ==
         "Effector memory CD8 T cells",
         ifelse(so@meta.data[[g.b]] == sel_cluster,
                "Proliferating effector memory CD8 T cells",
                "Non-proliferating effector memory CD8 T cells"),
         as.character(so$pruned.labels_singleR_fine))
```

# After splitting

```{r after, echo=F}
table(so$pruned.labels_singleR_fine.mod)
```

# Post-processing

Repeat the steps of Harmony post-processing (group cell types, abbreviate
labels, create color schemes), adding the two additional cell types derived
from splitting.

## Group and sort fine celltype labels

```{r fctype_groups, echo=TRUE}
so@misc$fctype_mod_group <- list()
so@misc$fctype_mod_group[["CD4+"]] <-
  c("Th17 cells", "Th2 cells", "Th1 cells",
    "Th1/Th17 cells", "Naive CD4 T cells", "T regulatory cells",
    "Follicular helper T cells", "Terminal effector CD4 T cells")
so@misc$fctype_mod_group[["CD8+"]] <-
  c("Naive CD8 T cells", "Central memory CD8 T cells",
    "Terminal effector CD8 T cells",
    "Proliferating effector memory CD8 T cells",
    "Non-proliferating effector memory CD8 T cells")
so@misc$fctype_mod_group[["Other"]] <-
  c("Gamma-delta T cells", "MAIT cells",
    "Natural killer cells", "Uncharacterized")

so$pruned.labels_singleR_fine.mod <-
  factor(so$pruned.labels_singleR_fine.mod,
         levels = c(so@misc$fctype_mod_group[["CD4+"]],
                    so@misc$fctype_mod_group[["CD8+"]],
                    so@misc$fctype_mod_group$Other))
so$pruned.labels_singleR_fine.mod <-
  droplevels(so$pruned.labels_singleR_fine.mod)
```

## Abbreviated labels

```{r abbrev_labels, echo=TRUE}
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Proliferating effector memory CD8 T cells"]] <- "Prolif. CD8+ EM"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Non-proliferating effector memory CD8 T cells"]] <- "Non-prolif. CD8+ EM"

so$pruned.labels_singleR_fine.mod.abbrev <- factor(
  as.vector(so@misc$pruned.labels_singleR_fine.abbrev[
              as.character(so$pruned.labels_singleR_fine.mod)]),
  levels = so@misc$pruned.labels_singleR_fine.abbrev[
              levels(so$pruned.labels_singleR_fine.mod)]
)
```

## Color schemes

```{r color_scheme_for_fine_labels}
so@misc$fctype_mod_cols <-
  c(so@misc$fctype_cols,
   "Proliferating effector memory CD8 T cells" = "#ffa9ff",  # Light Pink
   "Non-proliferating effector memory CD8 T cells" = "#ac2355") # Dark Pink
```

## Cell proportions permutations test

```{r cell_proportions_permutation_test, echo=TRUE}
so@misc$proptest_mod <- permutation_test(sc_utils(so),
                cluster_identity = "pruned.labels_singleR_fine.mod.abbrev",
                sample_1 = "Subclinical", sample_2 = "Hospitalized",
                sample_identity = "clinical_condition",
                n_permutations = 1000)
```

## Saving data

```{r, echo=TRUE}
saveRDS(so, paste0(var_dir, "splitted_CD8EM_so.rds"))
```

# Session info

```{r}
sessionInfo()
```
