---
title: "InDen T cells | Post-processing of Platypus vgm"
author: "Giorgio Gonnella"
date: "2024-10-14"
output:
  html_document:
    theme: sandstone
params:
  prjpath: "/srv/baia/inden/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=T)
.libPaths(c(params$libpath, .libPaths()))

library(Platypus)
library(dplyr)

analysis_root <- paste0(params$prjpath, "analysis-1/")

input_step_dir <- paste0(analysis_root, "steps/010.0.P.T.TCR_analysis.platypus/")
input_var_dir <- paste0(input_step_dir, "results/vars/")
if (!dir.exists(input_var_dir)) {
  stop("Input variable directory does not exist: ", input_var_dir)
}

this_step_dir <- paste0(analysis_root, "steps/010.0.P.T.TCR_analysis.platypus/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}

var_dir <- paste0(this_step_dir, "results/vars/")
dir.create(var_dir, showWarnings = FALSE, recursive = TRUE)

plots_dir <- paste0(this_step_dir, "results/plots/")
dir.create(plots_dir, showWarnings = FALSE, recursive = TRUE)

helpers_dir <- paste0(params$prjpath, "helpers/")
source(paste0(helpers_dir, "platypus_helpers.R"))
```

# Load data

```{r load_vgm}
vgm <- readRDS(file = paste0(input_var_dir, "vgm_raw.rds"))
```

# Post-processing

## Cleanup metadata columns

```{r cleanup_metadata_columns}
columns_to_remove <-
  setdiff(grep("_fine", colnames(vgm[["GEX"]]@meta.data), value=T),
          grep("pruned.*_fine", colnames(vgm[["GEX"]]@meta.data), value=T))
columns_to_remove <- c(columns_to_remove,
    setdiff(grep("_main", colnames(vgm[["GEX"]]@meta.data), value=T),
            grep("pruned.*_main", colnames(vgm[["GEX"]]@meta.data), value=T)))
vgm[["GEX"]]@meta.data <- vgm[["GEX"]]@meta.data[,
      !(colnames(vgm[["GEX"]]@meta.data) %in% columns_to_remove)]
```

## Fix GEX barcodes

The GEX barcodes are in the format (orig.barcode)-(sfx) e.g. AAAGCAAAGAACTGTA-1
which is the same as in the imported Seurat object, but the VDJ barcodes are in
the format (sample\_id)\_(orig.barcode) e.g. s9\_AAAGCAAAGAACTGTA - which is
correctly set as cell names in GEX but not as the barcode column.
Thus we will fix the barcode column to be the same as the names.

The VDJ\_avail and GEX\_avail columns are already correctly computed.

```{r fix_barcodes}
vgm[["GEX"]]$barcode = names(vgm[["GEX"]]$orig.ident)
```

## Add sample names

The sample column of VDJ is set to the same names as the sample column of GEX,
based on the sample\_id column and the sample\_id\_table stored in GEX.

In particular, sample are the first sample IDs used (which do not code the type
of condition) and sample\_name are the second set of IDs used (which start with
the first letter of the clinical condition, followed by a number, and exclude
the samples which were filtered out). Sample\_name is what is used in the
manuscript.

```{r create_vdj_sample_column}
index <- match(vgm[["VDJ"]]$sample_id,
               vgm[["GEX"]]@misc$sample_id_table$sample_id)
vgm[["VDJ"]]$sample <- factor(vgm[["GEX"]]@misc$sample_id_table$sample[index],
                              levels = levels(vgm[["GEX"]]$sample))
```

```{r transfer_sample_name_to_vdj}
vgm <- transfer_gex_to_vdj(vgm, "sample_name")
```

## Add clinical condition column

```{r create_sample_clinical_condition_table}
vgm[["GEX"]]@misc$sample_cond_table <-
  vgm[["GEX"]]@meta.data %>% group_by(sample) %>%
    summarise(clinical_condition = paste0(clinical_condition[[1]]))
```

```{r create vdj_clinical_condition_column}
index <- match(vgm[["VDJ"]]$sample,
               vgm[["GEX"]]@misc$sample_cond_table$sample)
vgm[["VDJ"]]$clinical_condition <-
  factor(vgm[["GEX"]]@misc$sample_cond_table$clinical_condition[index],
         levels = levels(vgm[["GEX"]]$clinical_condition))
rm(index)
vgm[["VDJ"]]$group_id <- vgm[["VDJ"]]$clinical_condition
```

## Select celltype and clusters metadata from GEX previous analysis

Copy the celltype annotations to a column "celltype" which is standard in the
Platypus examples, and copy the the column to the VDJ for which GEX is avail,
setting it to NA for all other (originally it is set to "T cell" for all VDJ).

```{r transfer_celltypes}
vgm[["GEX"]]$pruned.labels_singleR_fine.mod <-
  droplevels(vgm[["GEX"]]$pruned.labels_singleR_fine.mod)
vgm[["GEX"]]$celltype <- vgm[["GEX"]]$pruned.labels_singleR_fine.mod
vgm[["GEX"]]$celltype.abbrev <-
  vgm[["GEX"]]$pruned.labels_singleR_fine.mod.abbrev

vgm <- transfer_gex_to_vdj(vgm, "pruned.labels_singleR_fine.mod")
vgm <- transfer_gex_to_vdj(vgm, "pruned.labels_singleR_fine.mod.abbrev")
vgm[["VDJ"]]$celltype <- NA
vgm[["VDJ"]]$celltype <- vgm[["VDJ"]]$pruned.labels_singleR_fine.mod
vgm[["VDJ"]]$celltype.abbrev <-
  vgm[["VDJ"]]$pruned.labels_singleR_fine.mod.abbrev
# also transfer the unsplitted version:
vgm <- transfer_gex_to_vdj(vgm, "pruned.labels_singleR_fine")
vgm <- transfer_gex_to_vdj(vgm, "pruned.labels_singleR_fine.abbrev")
```

Transfer the cluster annotation at the selected clustering resolution to
the VDJ and copy it in the GEX and VDJ to the "seurat\_clusters" column,
which is used in the Platypus examples.

```{r transfer_clusters}
vgm[["GEX"]]$seurat_clusters <-
  vgm[["GEX"]]@meta.data[[vgm[["GEX"]]@misc$sel_res_col]]

vgm <- transfer_gex_to_vdj(vgm, vgm[["GEX"]]@misc$sel_res_col)
vgm[["VDJ"]]$seurat_clusters <- vgm[["VDJ"]][[vgm[["GEX"]]@misc$sel_res_col]]
```

## Filtering

### Plots before filtering

```{r cmp_n_cells}
gex_data <- data.frame(sample = factor(vgm[["GEX"]]$sample), library = "GEX",
                       celltype = factor(vgm[["GEX"]]$celltype.abbrev))
vdj_data <- data.frame(sample = factor(vgm[["VDJ"]]$sample), library = "VDJ",
                       celltype = factor(vgm[["VDJ"]]$celltype.abbrev))
combined_data <- bind_rows(gex_data, vdj_data)
plt <- ggplot(combined_data, aes(x = library, y = ..count.., fill=celltype)) +
  geom_bar(stat = "count") +
  facet_wrap(~sample,nrow=1)+
  xlab("Sample") + ylab("N.cells") +
  ggtitle("Number of cells") +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = vgm[["GEX"]]@misc$fctype_mod_cols)
ggsave(paste0(plots_dir, "before_filtering.n_cells_per_sample.pdf"), plt)
plt
```

```{r first_analysis_plots}
plt1 <- ggplot(data.frame(values=vgm[["GEX"]]$sample),
       aes(x=values, fill=values)) + geom_bar() +
       theme(legend.position="none") + xlab("Sample") + ylab("N.cells") +
       ggtitle("Number of cells in the GEX library") +
       scale_fill_manual(values = vgm[["GEX"]]@misc$sample_cols)
plt2 <- ggplot(data.frame(values=vgm[["VDJ"]][["sample"]]),
       aes(x=values, fill=values)) + geom_bar() +
       theme(legend.position="none") + xlab("Sample") + ylab("N.cells") +
       ggtitle("Number of cells in the VDJ library") +
       scale_fill_manual(values = vgm[["GEX"]]@misc$sample_cols)
plt <- grid.arrange(plt1, plt2, ncol=1)
plt
```

```{r plot_analysis_celltypes_before_filtering, fig.height=25, fig.width=8}
plt1 <- Seurat::DimPlot(vgm[["GEX"]],reduction = "umapharm",
                       group.by = "seurat_clusters",
                       cols = vgm[["GEX"]]@misc$cluster_cols)
plt2 <- Seurat::DimPlot(vgm[["GEX"]],reduction = "umapharm",
                       group.by = "celltype.abbrev",
                       cols = vgm[["GEX"]]@misc$fctype_mod_cols)
plt3 <- Seurat::DimPlot(vgm[["GEX"]],reduction = "umapharm",
                      group.by = "sample",
                      cols = vgm[["GEX"]]@misc$sample_cols)
plt4 <- Seurat::DimPlot(vgm[["GEX"]],reduction = "umapharm",
                       group.by = "VDJ_available")
plt <- grid.arrange(plt1, plt2, plt3, plt4, ncol=1)
print(plt)
rm(plt, plt1, plt2, plt3, plt4)
```

### Table GEX/VDJ available

```{r n_unavailable}
print("VDJ available for cells in GEX dataset:")
table(vgm[["GEX"]]$VDJ_available)

print("GEX available for cells in VDJ dataset:")
table(vgm[["VDJ"]]$GEX_available)

print("VDJ available for cells in GEX dataset, by sample:")
table(vgm[["GEX"]]$sample, vgm[["GEX"]]$VDJ_available)

print("GEX available for cells in VDJ dataset, by sample:")
table(vgm[["VDJ"]]$sample, vgm[["VDJ"]]$GEX_available)

print("VDJ available for cells in GEX dataset, by celltype:")
vgm[["GEX"]]$celltype <- droplevels(vgm[["GEX"]]$celltype)
vgm[["GEX"]]$celltype.abbrev <- droplevels(vgm[["GEX"]]$celltype.abbrev)
table(vgm[["GEX"]]$celltype.abbrev, vgm[["GEX"]]$VDJ_available)

print("VDJ available for cells in GEX dataset, by cluster:")
table(vgm[["GEX"]]$seurat_clusters, vgm[["GEX"]]$VDJ_available)
```

## Filter VJ sequence too long

A few sequences have an abnormal VJ sequence.
We will filter sequences with a VJ sequence longer than 30 amino acids.

```{r filter_vj_too_long}
vgm <- vgm_filter_vj_too_long(vgm, 30)
```

## Restrict to VDJ/GEX both available

```{r filter_unavailable}
vgm <- vgm_filter_unavailable(vgm)
```

## Plots after restricting to VDJ/GEX both available

```{r VDJ_avail_analysis_plots}
plt <- ggplot(data.frame(values=vgm[["GEX"]]$sample),
       aes(x=values, fill=values)) + geom_bar() +
       theme(legend.position="none") + xlab("Sample") + ylab("N.cells") +
       ggtitle("Number of cells in the GEX and VDJ library") +
       scale_fill_manual(values = vgm[["GEX"]]@misc$sample_cols)
ggsave(paste0(plots_dir, "after_vdj_avail_filter.n_cells_per_sample.pdf"), plt)
plt
```

### Aberrant n.of chains

```{r n_aberrant}
print("Missing VDJ chain: ")
table(subset(vgm[["VDJ"]], Nr_of_VDJ_chains == 0)$sample)
print("More than 1 VDJ chains: ")
table(subset(vgm[["VDJ"]], Nr_of_VDJ_chains > 1)$sample)
print("Missing VJ chain: ")
table(subset(vgm[["VDJ"]], Nr_of_VJ_chains == 0)$sample)
print("More than 1 VJ chains: ")
table(subset(vgm[["VDJ"]], Nr_of_VJ_chains > 1)$sample)
```

```{r}
total_counts <- table(vgm[["VDJ"]]$sample)
missing_vdj_counts <- table(subset(vgm[["VDJ"]], Nr_of_VDJ_chains == 0)$sample)
missing_vdj_percentages <- round((missing_vdj_counts /
                                  total_counts[names(missing_vdj_counts)]) * 100, 2)
more_than_1_vdj_counts <- table(subset(vgm[["VDJ"]], Nr_of_VDJ_chains > 1)$sample)
more_than_1_vdj_percentages <- round((more_than_1_vdj_counts /
                                      total_counts[names(more_than_1_vdj_counts)]) * 100, 2)
missing_vj_counts <- table(subset(vgm[["VDJ"]], Nr_of_VJ_chains == 0)$sample)
missing_vj_percentages <- round((missing_vj_counts /
                                 total_counts[names(missing_vj_counts)]) * 100, 2)
more_than_1_vj_counts <- table(subset(vgm[["VDJ"]], Nr_of_VJ_chains > 1)$sample)
more_than_1_vj_percentages <- round((more_than_1_vj_counts /
                                     total_counts[names(more_than_1_vj_counts)]) * 100, 2)
total_abnormal_counts <- missing_vdj_counts + more_than_1_vdj_counts + missing_vj_counts + more_than_1_vj_counts
total_abnormal_percentages <- round(total_abnormal_counts /
                                    total_counts[names(total_abnormal_counts)] * 100, 2)

results <- data.frame(
  sample = names(total_counts),
  missing_vdj_count = as.integer(missing_vdj_counts[names(total_counts)]),
  missing_vdj_percent = as.vector(missing_vdj_percentages[names(total_counts)]),
  more_than_1_vdj_count = as.integer(more_than_1_vdj_counts[names(total_counts)]),
  more_than_1_vdj_percent = as.vector(more_than_1_vdj_percentages[names(total_counts)]),
  missing_vj_count = as.integer(missing_vj_counts[names(total_counts)]),
  missing_vj_percent = as.vector(missing_vj_percentages[names(total_counts)]),
  more_than_1_vj_count = as.integer(more_than_1_vj_counts[names(total_counts)]),
  more_than_1_vj_percent = as.vector(more_than_1_vj_percentages[names(total_counts)]),
  total_abnormal_counts = as.integer(total_abnormal_counts[names(total_counts)]),
  total_abnormal_percent = as.vector(total_abnormal_percentages[names(total_counts)])
)

# Replace NAs with 0 for counts and percentages
results[is.na(results)] <- 0

# Print the table using knitr::kable
knitr::kable(results, caption = "Summary of VDJ and VJ Chain Abnormalities",
             col.names = c("Sample",
                             "Missing VDJ Count", "Missing VDJ Percent",
                             "More than 1 VDJ Count", "More than 1 VDJ Percent",
                             "Missing VJ Count", "Missing VJ Percent",
                             "More than 1 VJ Count", "More than 1 VJ Percent",
                             "Total abnormal Count", "Total abnormal Percent"))
```

```{r filter_aberrant}
vgm <- require_one_H_and_not_more_than_one_L(vgm)
```

Due to the large number of cells without VJ, we will create two different
datasets, one containing all sequences, one only sequences which also have the
VJ.

```{r cmp_n_cells_after_filter_aberrant}
gex_data <- data.frame(sample = factor(vgm[["GEX"]]$sample), library = "GEX",
                       celltype = factor(vgm[["GEX"]]$celltype.abbrev))
vdj_data <- data.frame(sample = factor(vgm[["VDJ"]]$sample), library = "VDJ",
                       celltype = factor(vgm[["VDJ"]]$celltype.abbrev))
combined_data <- bind_rows(gex_data, vdj_data)
combined_plot <- ggplot(combined_data, aes(x = library, y = ..count.., fill=celltype)) +
  geom_bar(stat = "count") +
  facet_wrap(~sample,nrow=1)+
  xlab("Sample") + ylab("N.cells") +
  ggtitle("Number of cells") +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = vgm[["GEX"]]@misc$fctype_mod_cols)
print(combined_plot)
ggsave(paste0(plots_dir, "after_aberrant_filter.n_cells_per_sample.pdf"), combined_plot)
```

### Table GEX/VDJ available after filtering aberrant n.of chains

```{r n_unavailable_after_filter_aberrant}
print("VDJ available for cells in GEX dataset:")
table(vgm[["GEX"]]$VDJ_available)

print("VDJ available for cells in GEX dataset, by sample:")
table(vgm[["GEX"]]$sample, vgm[["GEX"]]$VDJ_available)

print("VDJ available for cells in GEX dataset, by celltype:")
vgm[["GEX"]]$celltype <- droplevels(vgm[["GEX"]]$celltype)
vgm[["GEX"]]$celltype.abbrev <- droplevels(vgm[["GEX"]]$celltype.abbrev)
table(vgm[["GEX"]]$celltype.abbrev, vgm[["GEX"]]$VDJ_available)

print("VDJ available for cells in GEX dataset, by cluster:")
table(vgm[["GEX"]]$seurat_clusters, vgm[["GEX"]]$VDJ_available)
```

## Gene groups

```{r gene_groups}
vgm$VDJ$VJ_vgene_grp <-
  sapply(strsplit(vgm$VDJ$VJ_vgene, "-"), function(x) x[1])
vgm$VDJ$VDJ_vgene_grp <-
  sapply(strsplit(vgm$VDJ$VDJ_vgene, "-"), function(x) x[1])
vgm$VDJ$VDJ_jgene_locus <- sub("-.*", "", vgm$VDJ$VDJ_jgene)

vgm <- transfer_vdj_to_gex(vgm, c("VJ_vgene_grp", "VDJ_vgene_grp",
                                  "VDJ_jgene_locus", "VDJ_vgene",
                                  "VDJ_dgene", "VDJ_jgene", "VDJ_cgene",
                                  "VJ_vgene", "VJ_jgene", "VJ_cgene"))

# transform all empty strings to NA
for (col in c("VJ_vgene_grp", "VDJ_vgene_grp",
              "VDJ_jgene_locus", "VDJ_vgene",
              "VDJ_dgene", "VDJ_jgene", "VDJ_cgene",
              "VJ_vgene", "VJ_jgene", "VJ_cgene")) {
  vgm[["GEX"]]@meta.data[[col]] <- ifelse(vgm[["GEX"]]@meta.data[[col]] == "",
                                           NA, vgm[["GEX"]]@meta.data[[col]])
}
```

## Clonotypes

```{r}
vgm$VDJ <- VDJ_clonotype(VDJ=vgm$VDJ, clone.strategy = "cdr3.aa",
                         hierarchical = "single.chains", global.clonotype = T)
```

Compute for each cell the size of the clonotype to which it belongs.

```{r clonotype_size_vgm}
vgm <- cells_clonotype_size(vgm)
vgm <- compute_clonotype_size_bins(vgm)
```

Group clonotype sizes into bins for plotting and setup bin colors for plotting.

```{r clonotype_size_bins_vgm}
vgm$GEX@misc$clonotype_size_bin_cols <-
  c("1" = "#e6e6e6", "2-5" = "#ac9393", "6-20" = "#fc8d59", ">20" = "#ff0000")
```

## Saving Data

```{r save_results}
saveRDS(vgm, paste0(var_dir, "vgm.rds"))
```

## Session Info

```{r session_info, echo=FALSE}
sessionInfo()
```

