---
title: "InDen B cells | Split Plasmablasts"
author: "Giorgio Gonnella"
date: 2024-06-27
output:
  html_document:
    theme: sandstone
params:
  pfx: "/srv/baia/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=T)
.libPaths(c(params$libpath, .libPaths()))

library(Seurat)
library(scProportionTest)

prj_root <- paste0(params$pfx, "inden/")
analysis_root <- paste0(prj_root, "analysis-1/")

prev_step_dir <- paste0(analysis_root, "steps/104.0.P.B.integration.harmony/")
input_var_dir <- paste0(prev_step_dir, "results/vars/")
if (!dir.exists(input_var_dir)) {
  stop("Input variable directory does not exist: ", input_var_dir)
}

this_step_dir <- paste0(analysis_root, "steps/105.0.P.B.split_plasmablasts/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}
var_dir <- paste0(this_step_dir, "results/vars/")
dir.create(var_dir, showWarnings = FALSE, recursive = TRUE)

```

```{r load, include = FALSE}
so <- readRDS(file = paste0(input_var_dir, "integrated_so.rds"))
```

The purpose of this step is to divide the plasmablasts according to
their unsupervised clustering. The proliferating cluster cells will be separated
from the other cells, and called, respectively, "Proliferating plasmablasts"
and "Non-proliferating plasmablasts".

# Before splitting

```{r before}
table(so$pruned.labels_singleR_fine)
```

# Splitting

```{r splitting, echo=TRUE}
g.b = so@misc$sel_res_col

# Select the cluster with the lowest G1 proportion
G1_cluster_p =
  prop.table(table(so$Phase, so@meta.data[[g.b]]), margin=2)["G1",]
sel_cluster = names(G1_cluster_p[which.min(G1_cluster_p)])[[1]]
print(paste0("Proliferating cells cluster: ", sel_cluster))

so$pruned.labels_singleR_fine.mod <- so$pruned.labels_singleR_fine
so$pruned.labels_singleR_fine.mod <-
  ifelse(so$pruned.labels_singleR_fine ==
         "Plasmablasts",
         ifelse(so@meta.data[[g.b]] == sel_cluster,
                "Proliferating plasmablasts",
                "Non-proliferating plasmablasts"),
         as.character(so$pruned.labels_singleR_fine))
```

# After splitting

```{r after}
table(so$pruned.labels_singleR_fine.mod)
```

# Post-processing

Repeat the steps of Harmony post-processing (group cell types, abbreviate
labels, create color schemes), adding the two additional cell types derived
from splitting.

## Sort fine celltype labels

```{r fctype_sort, echo=TRUE}
order <- c("Naive B cells", "Non-switched memory B cells", "Switched memory B cells",
    "Non-proliferating plasmablasts", "Proliferating plasmablasts", "Exhausted B cells")
so$pruned.labels_singleR_fine.mod <-
  factor(so$pruned.labels_singleR_fine.mod, levels = order)
```

## Abbreviated labels

```{r abbrev_labels, echo=TRUE}
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Proliferating plasmablasts"]] <- "Prol. plasmab."
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Non-proliferating plasmablasts"]] <- "Non-prol. plasmab."

so$pruned.labels_singleR_fine.mod.abbrev <- factor(
  as.vector(so@misc$pruned.labels_singleR_fine.abbrev[as.character(so$pruned.labels_singleR_fine.mod)]),
  levels = so@misc$pruned.labels_singleR_fine.abbrev[levels(so$pruned.labels_singleR_fine.mod)]
)
```

## Color schemes
```{r color_scheme_for_fine_labels}
so@misc$fctype_mod_cols <-
  c(so@misc$fctype_cols,
   "Proliferating plasmablasts" = "#46531a",
   "Non-proliferating plasmablasts" = "#238b45")
```

## Cell proportions permutations test

```{r cell_proportions_permutation_test, echo=TRUE}
so@misc$proptest_mod <- permutation_test(sc_utils(so),
                cluster_identity = "pruned.labels_singleR_fine.mod.abbrev",
                sample_1 = "Subclinical", sample_2 = "Hospitalized",
                sample_identity = "clinical_condition",
                n_permutations = 1000)
```

## Saving data

```{r, echo=TRUE}
saveRDS(so, file = paste0(var_dir, "splitted_so.rds"))
```

# Session info

```{r}
sessionInfo()
```
