---
title: "InDen T cells | Results after Splitting CD8+ EM cells"
author: "Giorgio Gonnella"
date: 2024-04-23
output:
  html_document:
    theme: sandstone
params:
  pfx: "/srv/baia/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=F)
.libPaths(c(params$libpath, .libPaths()))

library(Seurat)
library(ggplot2)

prj_root <- paste0(params$pfx, "inden/")
analysis_root <- paste0(prj_root, "analysis-1/")

input_step_dir <- paste0(analysis_root, "steps/008.0.P.T.split_CD8EM/")
input_var_dir <- paste0(input_step_dir, "results/vars/")
if (!dir.exists(input_var_dir)) {
  stop("Input variable directory does not exist: ", input_var_dir)
}

this_step_dir <- paste0(analysis_root, "steps/008.0.V.T.splitted_CD8EM/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}

tables_outdir <- paste0(this_step_dir, "results/tables/")
plots_outdir <- paste0(this_step_dir, "results/plots/")
dir.create(tables_outdir, recursive = T, showWarnings = F)
dir.create(plots_outdir, recursive = T, showWarnings = F)

helpers_dir <- paste0(prj_root, "helpers/")
source(paste0(helpers_dir, "combined_proportion_plot.R"))
source(paste0(helpers_dir, "dimplot_helpers.R"))
```

```{r load, echo = FALSE}
so <- readRDS(paste0(input_var_dir, "splitted_CD8EM_so.rds"))
```

```{r cond_subsets, echo=F}
cond_subset <- list()
for (cond in levels(so$clinical_condition)) {
  cond_subset[[cond]] <- subset(so,
                                subset = clinical_condition == cond)
  cond_subset[[cond]]$sample <- droplevels(cond_subset[[cond]]$sample)
}
```

# Clinical condition table

```{r clinical_condition, results="asis", echo=F}
print(knitr::kable(table(
        so$clinical_condition)))
cat("\n\n")
print(knitr::kable(as.data.frame.matrix(table(so$sample,
                   so$clinical_condition))))
```

# Cell types

## By clinical condition

```{r singleR_fine.mod_umap, fig.width=12, fig.height=6, echo=F}
plt <- DimPlot(object = so, reduction = "umapharm",
        group.by = "pruned.labels_singleR_fine.mod",
        cols = so@misc$fctype_mod_cols, label = F, repel = TRUE,
        split.by = "clinical_condition")
plt <- plt + ggtitle("T cells subtype assignments by SingleR")
plt <- plt + labs(x = "UMAP Harmony 1", y = "UMAP Harmony 2")
fname = paste0(plots_outdir, "umap.by_cond.singleR_fine.mod.svg")
ggsave(fname, plot = plt, device = "svg")
plt
```

### Absolute counts

```{r table_cond_per_fine_celltypes_abs_file, echo=F}
tab <- as.data.frame.matrix(table(
  so$pruned.labels_singleR_fine.mod,
  so$clinical_condition))
write.table(tab,
  file = paste0(tables_outdir, "table_cond_per_fine_celltypes_abs.tsv"),
  sep = "\t")
knitr::kable(tab)
```

```{r}
cell_type_counts <- so@meta.data %>%
  group_by(pruned.labels_singleR_fine.mod, clinical_condition) %>%
  summarise(count = n()) %>%
  ungroup()

cell_type_counts$pruned.labels_singleR_fine.mod <-
  factor(cell_type_counts$pruned.labels_singleR_fine.mod,
         levels = levels(so$pruned.labels_singleR_fine.mod))
```

```{r}
plt1 <- ggplot(cell_type_counts, aes(x = count,
                                     y = pruned.labels_singleR_fine.mod,
                                     fill = clinical_condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(x = "Count", y = "Cell Type",
       title = "Frequency of Cell Types by Condition") +
  scale_fill_manual(values = so@misc$cond_cols) +
  theme(axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.title = element_text(size = 14),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))
print(plt1)
```

### Relative proportions

```{r table_fine_celltypes_per_cond, echo=F}
knitr::kable(as.data.frame.matrix(round(prop.table(table(so$clinical_condition,
  so$pruned.labels_singleR_fine.mod), margin = 2)*100, digits = 2)))
```

```{r table_cond_per_fine_celltypes_file, echo=F}
tab <- as.data.frame.matrix(round(prop.table(table(
  so$pruned.labels_singleR_fine.mod,
  so$clinical_condition), margin = 2)*100, digits = 2))
write.table(tab,
  file = paste0(tables_outdir, "table_cond_per_fine_celltypes.tsv"),
  sep = "\t")
knitr::kable(tab)
```

```{r}
total_counts <- cell_type_counts %>%
  group_by(clinical_condition) %>%
  summarise(total = sum(count))

relative_counts <- cell_type_counts %>%
  left_join(total_counts, by = "clinical_condition") %>%
  mutate(relative_count = count / total) %>%
  select(clinical_condition, pruned.labels_singleR_fine.mod, relative_count)
```

```{r}
plt2 <- ggplot(relative_counts, aes(x = relative_count,
                                    y = pruned.labels_singleR_fine.mod,
                                    fill = clinical_condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(x = "Proportion", y = "Cell Type",
       title = "Relative Frequency of Cell Types by Condition") +
  scale_fill_manual(values = so@misc$cond_cols) +
  theme(axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.title = element_text(size = 14),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))
print(plt2)
```

```{r celltype_df_f_prep, echo=F}
celltype_df_f <- list()
for (cond in levels(so$clinical_condition)) {
  celltype_df_f[[cond]] <-
    as.data.frame(round(prop.table(table(cond_subset[[cond]]$sample,
                    cond_subset[[cond]]$pruned.labels_singleR_fine.mod),
                      margin = 2)*100, digits = 2))
  colnames(celltype_df_f[[cond]]) <- c("sample_id", "cell_type", "percentage")
  # remove sample which are not in the current condition
  celltype_df_f[[cond]] <-
    celltype_df_f[[cond]][celltype_df_f[[cond]]$sample_id %in%
                            unique(cond_subset[[cond]]$sample),]
  celltype_df_f[[cond]]$sample_id <-
    factor(celltype_df_f[[cond]]$sample_id,
           levels = levels(cond_subset[[cond]]$sample))
  #droplevels(celltype_df_f[[cond]]$sample_id)
}
```

```{r celltype_df_f_plots, fig.width=12, fig.height=4, echo=F}
for (cond in levels(so$clinical_condition)) {
  print(ggplot(celltype_df_f[[cond]], aes(x = cell_type, y = percentage))+
    geom_bar(stat = "identity", aes(fill = sample_id))+
    ggtitle(cond)+
    scale_fill_manual(values = so@misc$sample_cols)+
    theme_bw()+
    coord_flip()+
    scale_x_discrete(limits = rev(levels(celltype_df_f[[cond]]$cell_type))))
}
```

```{r celltype_df_fr_prep, echo=F}
celltype_df_fr <- list()
for (cond in levels(so$clinical_condition)) {
  celltype_df_fr[[cond]] <- as.data.frame(round(prop.table(table(
        cond_subset[[cond]]$pruned.labels_singleR_fine.mod,
        cond_subset[[cond]]$sample),
          margin = 2)*100, digits = 2))
  colnames(celltype_df_fr[[cond]]) <- c("cell_type", "sample_id", "percentage")
  # remove sample which are not in the current condition
  celltype_df_fr[[cond]] <-
    celltype_df_fr[[cond]][celltype_df_fr[[cond]]$sample_id %in%
                             unique(cond_subset[[cond]]$sample),]
  celltype_df_fr[[cond]]$sample_id <-
    factor(celltype_df_fr[[cond]]$sample_id,
          levels = levels(cond_subset[[cond]]$sample))
  #droplevels(celltype_df_fr[[cond]]$sample_id)
}
```

```{r celltype_df_fr_plots, fig.width=12, fig.height=5, echo=F}
for (cond in levels(so$clinical_condition)) {
  print(ggplot(celltype_df_fr[[cond]], aes(x = sample_id, y = percentage))+
    geom_bar(stat = "identity", aes(fill = cell_type))+
    ggtitle(cond)+
    scale_fill_manual(values = so@misc$fctype_mod_cols)+
    theme_bw()+
    coord_flip()+
    scale_x_discrete(limits = rev(levels(celltype_df_fr[[cond]]$sample_id)))) +
    theme(
      plot.margin = margin(t = 50, r = 120, b = 50, l = 10)
    )
}
```

### Proportion test

```{r proportion_plot, fig.width=8, fig.height=8, echo=F}
plt <- grouped_combined_permutation_test_plot(so, so@misc$proptest_mod,
            "pruned.labels_singleR_fine.mod.abbrev", "cell_group",
            "clinical_condition", "Clinical Condition", so@misc$cond_cols,
            so@misc$ctg_backcols)

fname <- paste0(plots_outdir, "singleR_proportion_test.svg")
ggsave(fname, plot = plt, device = "svg")
plt
```

```{r prop_test_results, echo=F}
cell_order <- c("Th17", "Th2", "Th1", "Th1/Th17",
                "CD4+ Naive", "Tregs", "Tfh",
                "CD4+ TEMRA", "CD8+ Naive", "CD8+ CM",
                "CD8+ TEMRA", "Prolif. CD8+ EM",
                "Non-prolif. CD8+ EM", "GD", "MAIT")

results <- so@misc$proptest_mod@results$permutation %>%
  mutate(clusters =
         factor(so@misc$proptest_mod@results$permutation$clusters,
                levels = cell_order)) %>%
  arrange(clusters)
write.table(results,
            file = paste0(tables_outdir, "singleR_proportion_test.tsv"),
            sep = "\t", row.names = F)
knitr::kable(results)
```

## By sample

```{r umap_singleR_by_sample_fine, fig.width=12, fig.height=12, echo=F}
DimPlot(object = so, reduction = "umapharm",
        group.by = "pruned.labels_singleR_fine.mod",
        cols = so@misc$fctype_mod_cols, label = F, repel = TRUE,
        split.by = "sample", ncol = 3) +
  NoLegend() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "orange1") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "orange1") +
  ggtitle("UMAP plot, colored by cell type, splitted by sample")
```

```{r table_fine_celltypes_per_sample, echo=F}
knitr::kable(as.data.frame.matrix(round(prop.table(table(so$sample,
  so$pruned.labels_singleR_fine.mod), margin = 2)*100, digits = 2)))
```

```{r plot_fine_celltypes_per_sample, fig.width=12, fig.height=6, echo=F}
celltype_df_f <- as.data.frame(round(prop.table(table(so$sample,
  so$pruned.labels_singleR_fine.mod), margin = 2)*100, digits = 2))
colnames(celltype_df_f) <- c("sample_id", "cell_type", "percentage")
ggplot(celltype_df_f, aes(x = cell_type, y = percentage))+
  geom_bar(stat = "identity", aes(fill = sample_id))+
  scale_fill_manual(values = so@misc$sample_cols)+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(celltype_df_f$cell_type)))
```

```{r table_samples_per_fine_celltype, echo=F}
knitr::kable(as.data.frame.matrix(round(prop.table(
    table(so$pruned.labels_singleR_fine.mod,
          so$sample), margin = 2)*100, digits = 2)))
```

```{r plot_samples_per_fine_celltypes, fig.width=12, echo=F}
celltype_df_fr <- as.data.frame(round(prop.table(
      table(so$pruned.labels_singleR_fine.mod,
            so$sample), margin = 2)*100, digits = 2))
colnames(celltype_df_fr) <- c("cell_type", "sample_id", "percentage")
ggplot(celltype_df_fr, aes(x = sample_id, y = percentage))+
  geom_bar(stat = "identity", aes(fill = cell_type))+
  scale_fill_manual(values = so@misc$fctype_mod_cols)+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(celltype_df_fr$sample_id)))
```

## CD8 EM cells

```{r, fig.width=8, fig.height=5, echo=F}
celltype1 <- "Proliferating effector memory CD8 T cells"
celltype2 <- "Non-proliferating effector memory CD8 T cells"
plt <- two_groups_dimplot(so, celltype1, celltype2,
                          "pruned.labels_singleR_fine.mod",
                          group_names = c("Proliferating", "Non-prolif."),
                          split_by = "clinical_condition")
fname = paste0(plots_outdir, "umap.by_cond.effectorCD8_splitted.svg")
ggsave(fname, plot = plt, device = "svg", width = 8, height = 5)
plt
```

```{r single_cell_types_umap_selected, echo=F}
celltype <- "Effector memory CD8 T cells"
highl <- WhichCells(so,
                      expression = pruned.labels_singleR_fine.mod == celltype)
plt <- DimPlot(so, reduction = "umapharm",
          cells.highlight = highl, cols.highlight = "darkblue",
          split.by = "clinical_condition",
          cols = "grey") + ggtitle(celltype) +
          theme(legend.position = "none")
fname = paste0(plots_outdir, "umap.by_cond.effectorCD8.svg")
ggsave(fname, plot = plt, device = "svg")
```

## UMAP highlighting single cell types

```{r single_cell_types_umap, fig.width=12, fig.height=6, echo=F}
for (celltype in unique(so$pruned.labels_singleR_fine.mod)) {
  hl <- WhichCells(so, expression = pruned.labels_singleR_fine.mod == celltype)
  plt <- DimPlot(so, reduction = "umapharm",
          cells.highlight = hl, cols.highlight = "darkblue",
          split.by = "clinical_condition", cols = "grey") + ggtitle(celltype) +
        theme(legend.position = "none")
  celltype_sanitized <- gsub("[^[:alnum:]]", "_", celltype)
  fname <- paste0(plots_outdir, "umap.by_cond.", celltype_sanitized, ".svg")
  ggsave(fname, plot = plt, device = "svg")
  print(plt)
}
```

## Phase

```{r phase, echo=F, fig.width=12}
plt <- DimPlot(object = so, reduction = "umapharm",
        group.by = "Phase", split.by = "clinical_condition",
        cols = so@misc$phase_cols) +
  ggtitle("UMAP plot, colored by phase")
fname = paste0(plots_outdir, "umap.by_cond.phase.svg")
ggsave(fname, plot = plt, device = "svg")
plt
```

### Phase markers

```{r phase_markers1, fig.height=5, fig.width=12, echo=F}
FeaturePlot(so, features = "PCNA",
            reduction = "umapharm", split.by = "clinical_condition",
            shape.by="Phase")
```

```{r phase_markers2, fig.height=5, fig.width=12, echo=F}
FeaturePlot(so, features = "TOP2A",
            reduction = "umapharm", split.by = "clinical_condition",
            shape.by="Phase")
```

```{r phase_markers3, fig.height=5, fig.width=12, echo=F}
FeaturePlot(so, features = "MCM6",
            reduction = "umapharm", split.by = "clinical_condition",
            shape.by="Phase")
```

```{r phase_markers4, fig.height=5, fig.width=12, echo=F}
plt <- FeaturePlot(so, features = "MKI67",
            reduction = "umapharm", split.by = "clinical_condition",
            shape.by="Phase")
fname = paste0(plots_outdir, "phase_marker.MKI67.svg")
ggsave(fname, plot = plt, device = "svg")
plt
```

### Celltype vs phase

#### Absolute counts

```{r phase_vs_fine_celltype_tables, results="asis", echo=F}
print(knitr::kable(as.data.frame.matrix(
      table(so$pruned.labels_singleR_fine.mod,
            so$Phase))))
print(knitr::kable(as.data.frame.matrix(addmargins(round(prop.table(
      table(so$pruned.labels_singleR_fine.mod,
            so$Phase),
      margin = 1)*100, digits = 2), margin = 2))))
print(knitr::kable(as.data.frame.matrix(addmargins(round(prop.table(
      table(so$Phase,
            so$pruned.labels_singleR_fine.mod),
      margin = 1)*100, digits = 2), margin = 2))))
```

#### Relative proportions

```{r phase_vs_fine_celltype_plots1, fig.width=12, fig.height=6, echo=F}
df <- as.data.frame(round(prop.table(table(
        so$pruned.labels_singleR_fine.mod,
        so$Phase), margin = 2)*100,
        digits = 2))
colnames(df) <- c("cell_type", "phase", "percentage")
ggplot(df, aes(x = phase, y = percentage))+
  geom_bar(stat = "identity", aes(fill = cell_type))+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(df$phase))) +
  scale_fill_manual(values = so@misc$fctype_mod_cols)
```

```{r phase_vs_fine_celltype_plots2, fig.width=12, fig.height=6, echo=F}
df2 <- as.data.frame(round(prop.table(table(
         so$Phase,
         so$pruned.labels_singleR_fine.mod), margin = 2)*100, digits = 2))
colnames(df2) <- c("phase", "cell_type", "percentage")
ggplot(df2, aes(x = cell_type, y = percentage))+
  geom_bar(stat = "identity", aes(fill = phase))+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(df2$cell_type))) +
  scale_fill_manual(values = so@misc$phase_cols)
```

## Clinical condition and unsupervised clustering

```{r selected_clustering_resolution, echo=F}
print(paste0("Selected clustering resolution: ",
             so@misc$sel_res))
```

```{r, echo=F}
knitr::kable(table(so$clinical_condition,
      so@meta.data[[so@misc$sel_res_col]], exclude = NULL))
```

```{r umap_clustering, fig.align="center", fig.width=12, fig.height=6, echo=F}
DimPlot(object = so, reduction = "umapharm",
        group.by = so@misc$sel_res_col, label = TRUE,
        cols = so@misc$cluster_cols,
        split.by = "clinical_condition")
```

### Sample contributions to clusters

```{r sample_vs_cl, echo=FALSE, results="asis"}
sample_vs_cl <- list()
for (cond in levels(so$clinical_condition)) {
  subset_table <- table(cond_subset[[cond]]@meta.data[[so@misc$sel_res_col]],
                        cond_subset[[cond]]$sample)
  cat(paste0("*Condition: ", cond, "*\n\n"))
  print(knitr::kable(subset_table))
  cat("\n\n")
  sample_vs_cl[[cond]] <-
    as.data.frame(round(prop.table(subset_table, margin=1)*100, digits = 2))
  colnames(sample_vs_cl[[cond]])[1:2] <- c("Cluster", "SampleID")
  levels(sample_vs_cl[[cond]]$Cluster) <-
    paste("Cl", levels(sample_vs_cl[[cond]]$Cluster), sep = "_")
}
```

```{r sample_vs_cl_plots, fig.width=12, fig.height=6, echo=FALSE}
for (cond in levels(so$clinical_condition)) {
  print(ggplot(sample_vs_cl[[cond]], aes(x = Cluster, y = Freq, fill = SampleID))+
    ggtitle(cond)+
    geom_bar(stat = "identity", alpha = .98)+
    coord_flip()+
    theme_minimal()+
    scale_fill_manual(values = so@misc$sample_cols))
}
```

```{r cl_vs_sample, echo=FALSE, results="asis"}
cl_vs_sample <- list()
for (cond in levels(so$clinical_condition)) {
  subset_table <- table(cond_subset[[cond]]$sample,
                        cond_subset[[cond]]@meta.data[[so@misc$sel_res_col]])
  cat(paste0("*Condition: ", cond, "*\n\n"))
  print(knitr::kable(subset_table))
  cat("\n\n")
  cl_vs_sample[[cond]] <-
    as.data.frame(round(prop.table(subset_table, margin=1)*100, digits = 2))
  colnames(cl_vs_sample[[cond]])[1:2] <- c("SampleID", "Cluster")
  levels(cl_vs_sample[[cond]]$Cluster) <-
    paste("Cl", levels(cl_vs_sample[[cond]]$Cluster), sep = "_")
}
```

```{r cl_vs_sample_plots, fig.width=12, fig.height=6, echo=FALSE}
cols <- so@misc$cluster_cols
names(cols) <- paste0("Cl_", names(so@misc$cluster_cols))
for (cond in levels(so$clinical_condition)) {
  print(ggplot(cl_vs_sample[[cond]],
               aes(x = SampleID, y = Freq, fill = Cluster))+
    ggtitle(cond)+
    geom_bar(stat = "identity", alpha = .98)+
    coord_flip()+
    theme_minimal()+
    scale_fill_manual(values = cols))
}
```

### Cell types by cluster

SingleR assignments by cluster.

```{r singleR_vs_clustering_tables, results="asis", echo=F}
print(knitr::kable(as.data.frame.matrix(
      table(so@meta.data[[so@misc$sel_res_col]],
            so$pruned.labels_singleR_fine.mod))))
print(knitr::kable(as.data.frame.matrix(addmargins(round(prop.table(
      table(so@meta.data[[so@misc$sel_res_col]],
            so$pruned.labels_singleR_fine.mod),
      margin = 1)*100, digits = 2), margin = 2))))
print(knitr::kable(as.data.frame.matrix(addmargins(round(prop.table(
      table(so$pruned.labels_singleR_fine.mod,
            so@meta.data[[so@misc$sel_res_col]]),
      margin = 1)*100, digits = 2), margin = 2))))
```

```{r singleR_vs_clustering_plots1, fig.width=12, fig.height=6, echo=F}
df <- as.data.frame(round(prop.table(table(so@meta.data[[so@misc$sel_res_col]],
        so$pruned.labels_singleR_fine.mod), margin = 2)*100,
        digits = 2))
colnames(df) <- c("cluster", "cell_type", "percentage")
ggplot(df, aes(x = cell_type, y = percentage))+
  geom_bar(stat = "identity", aes(fill = cluster))+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(df$cell_type))) +
  scale_fill_manual(values = so@misc$cluster_cols)
```

```{r singleR_vs_clustering_plots2, fig.width=12, fig.height=6, echo=F}
df2 <- as.data.frame(round(prop.table(table(
         so$pruned.labels_singleR_fine.mod,
         so@meta.data[[so@misc$sel_res_col]]), margin = 2)*100, digits = 2))
colnames(df2) <- c("cell_type", "cluster", "percentage")
ggplot(df2, aes(x = cluster, y = percentage))+
  geom_bar(stat = "identity", aes(fill = cell_type))+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(df2$cluster))) +
  scale_fill_manual(values = so@misc$fctype_mod_cols)
```

## Session info

```{r, echo=F}
sessionInfo()
```
