---
title: "InDen B cells | Clustering resolutions comparison"
author: "Giorgio Gonnella"
date: 2024-06-27
output:
  html_document:
    theme: sandstone
params:
  prjpath: "/srv/baia/prj/subclinical_dengue_scrnaseq/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=F)
.libPaths(c(params$libpath, .libPaths()))

library(Seurat)
library(viridis)
library(ggplot2)
library(RColorBrewer)
library(dplyr)
library(mclust)
library(clustree)
library(qualpalr)

analysis_root <- paste0(params$prjpath, "analysis-1/")

input_step_dir <- paste0(analysis_root, "steps/104.0.P.B.integration.harmony/")
var_dir <- paste0(input_step_dir, "results/vars/")
```

```{r load, echo=F}
so <- readRDS(file = paste0(var_dir, "integrated_so.rds"))
```

```{r clusters_palette}
cl_cols = unique(c(qualpal(99, colorspace="pretty")$hex,
            qualpal(99, colorspace="rainbow")$hex))
```

```{r resolution_lists, echo=F}
resolutions = grep("^harmsnn_res", colnames(so@meta.data),
                   value = TRUE)
# limit resolutions to consider for some of the plots
max_n_clusters = 60
n_clusters <- data.frame(Resolution = numeric(), N_Clusters = numeric(),
                         stringsAsFactors = FALSE)
for(g.b in resolutions){
  res <- gsub("harmsnn_res.", "", g.b)
  res_n_clusters = nrow(unique(so@meta.data[g.b]))
  n_clusters <- rbind(n_clusters,
        data.frame(Resolution = res,
                   N_Clusters = res_n_clusters))
}
res_l <- n_clusters$Resolution[n_clusters$N_Clusters <= max_n_clusters]
```

# Comparison of clustering at different resolutions

## Number of clusters

```{r table_n_clusters}
knitr::kable(n_clusters)
```

## Plots

```{r umap_clustering, fig.align="center", fig.width=12, fig.height=6, echo=F}
for (g.b in resolutions) {
  res <- gsub("harmsnn_res.", "", g.b)
  print(DimPlot(object = so, reduction = "umapharm",
          group.by = g.b, label = TRUE, cols = cl_cols))
  print(VlnPlot(object = so, features = c("nCount_RNA",
         "nFeature_RNA", "percent.mito"), group.by = g.b, cols = cl_cols))
  sample_vs_cl <- as.data.frame(
    round(prop.table(table(so@meta.data[[g.b]],
                           so$sample), margin = 1)*100,
          digits = 2)
  )
  colnames(sample_vs_cl)[1:2] <- c("Seurat_Cl", "Sampleid")
  levels(sample_vs_cl$Seurat_Cl) <-
    paste("Cl", levels(sample_vs_cl$Seurat_Cl), sep = "_")
  print(ggplot(sample_vs_cl, aes(x = Seurat_Cl, y = Freq, fill = Sampleid))+
    geom_bar(stat = "identity", alpha = .98)+
    coord_flip()+ theme_minimal()+
    scale_fill_manual(values = so@misc$sample_cols))
  # horizontal line to separate resolutions better in report
  print(ggplot() + geom_hline(yintercept = 0.5) + theme_void())
}
```

### Clustering tree

```{r clustree, fig.align="center", fig.width=12, fig.height=6, echo=F}
clustree(so, prefix = "harmsnn_res.")
```

## Cell types by cluster

### Adjusted Rand Index

We used the Adjusted Rand Index, computed using the mclust library, to
compare two partitions, in this case the single R assignments and the
unsupervised clustering. The value is zero for two random partitions and
one for perfect agreement betweeen the partitions.

```{r adj_rand_idx, echo=F}
ari <- data.frame(Resolution = numeric(), ARI_Value_fine = numeric(),
                  ARI_Value_main = numeric(), stringsAsFactors = FALSE)
for (g.b in resolutions) {
  ari_value_f <- adjustedRandIndex(so$pruned.labels_singleR_fine,
                                  so@meta.data[[g.b]])
  ari_value_m <- adjustedRandIndex(so$pruned.labels_singleR_main,
                                  so@meta.data[[g.b]])
  res <- gsub("harmsnn_res.", "", g.b)
  ari <- rbind(ari, data.frame(Resolution = res, ARI_Value_fine = ari_value_f,
                               ARI_Value_main = ari_value_m))
}
knitr::kable(ari)
```

### Weighted average cluster purity

We computed the cluster purity of each cluster as number of cells in the
most frequent cell type assignment divided by the total number of cells
of the cluster. We then computed an average of the cluster purity values
for all cluster, weighted by the cluster size. Additionally we compute
the percentage of clusters with a purity of at least 95%, 80%, 65% and
50% respectively.

```{r cluster_purity, echo=F}
purity <- data.frame(Resolution = numeric(),
                     Purity = numeric(),
                     perc_Clusters_Purity50 = numeric(),
                     perc_Clusters_Purity65 = numeric(),
                     perc_Clusters_Purity80 = numeric(),
                     perc_Clusters_Purity95 = numeric(),
                     stringsAsFactors = FALSE)

for (g.b in resolutions) {
  cluster_cell_types <- data.frame(
    Cluster = so@meta.data[[g.b]],
    Cell_Type_f = so$pruned.labels_singleR_fine
  )

  cluster_purity_details <- cluster_cell_types %>%
    group_by(Cluster) %>%
    count(Cell_Type_f) %>%
    mutate(TotalCells = sum(n)) %>%
    arrange(desc(n)) %>%
    slice(1) %>%
    ungroup() %>%
    mutate(Purity = n / TotalCells)

  weighted_average_purity <- sum(cluster_purity_details$Purity *
                      cluster_purity_details$TotalCells) /
                      sum(cluster_purity_details$TotalCells)

  n_clusters_purity50 <- sum(cluster_purity_details$Purity >= 0.50)
  n_clusters_purity65 <- sum(cluster_purity_details$Purity >= 0.65)
  n_clusters_purity80 <- sum(cluster_purity_details$Purity >= 0.80)
  n_clusters_purity95 <- sum(cluster_purity_details$Purity >= 0.95)

  perc_Clusters_Purity50 <- n_clusters_purity50 / nrow(cluster_purity_details) * 100
  perc_Clusters_Purity65 <- n_clusters_purity65 / nrow(cluster_purity_details) * 100
  perc_Clusters_Purity80 <- n_clusters_purity80 / nrow(cluster_purity_details) * 100
  perc_Clusters_Purity95 <- n_clusters_purity95 / nrow(cluster_purity_details) * 100

  res <- gsub("harmsnn_res.", "", g.b)
  purity <- rbind(purity, data.frame(Resolution = res,
                                     Purity_f = weighted_average_purity,
                                     perc_Clusters_Purity50 = perc_Clusters_Purity50,
                                     perc_Clusters_Purity65 = perc_Clusters_Purity65,
                                     perc_Clusters_Purity80 = perc_Clusters_Purity80,
                                     perc_Clusters_Purity95 = perc_Clusters_Purity95
                                     ))
}
knitr::kable(purity)
```

## Cell proportions by cluster

### Tables

```{r singleR_vs_clust_tab, fig.width=12, fig.height=6, results="asis", echo=F}
for (g.b in resolutions) {
  res = gsub("harmsnn_res.", "", g.b)
  cat("### Resolution : ", res, "\n")
  print(knitr::kable(as.data.frame.matrix(table(so@meta.data[[g.b]], so$pruned.labels_singleR_fine))))
  print(knitr::kable(as.data.frame.matrix(addmargins(round(prop.table(table(so@meta.data[[g.b]], so$pruned.labels_singleR_fine), margin = 1)*100, digits = 2), margin = 2))))
  print(knitr::kable(as.data.frame.matrix(addmargins(round(prop.table(table(so$pruned.labels_singleR_fine, so@meta.data[[g.b]]), margin = 1)*100, digits = 2), margin = 2))))
}
```

### Plots

SingleR assignments by cluster at different clustering resolutions (same data as above visualized as barplots).

```{r singleR_vs_clustering_plots, fig.width=12, fig.height=6, echo=F}
for (g.b in resolutions) {
  res = gsub("harmsnn_res.", "", g.b)
  cat("Resolution : ", res, "\n")
  df <- as.data.frame(round(prop.table(table(so@meta.data[[g.b]], so$pruned.labels_singleR_fine), margin = 2)*100, digits = 2))
  colnames(df) <- c("cluster", "cell_type", "percentage")
  print(ggplot(df, aes(x = cell_type, y = percentage))+
  geom_bar(stat = "identity", aes(fill = cluster))+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(df$cell_type))))
  df2 <- as.data.frame(round(prop.table(table(so$pruned.labels_singleR_fine, so@meta.data[[g.b]]), margin = 2)*100, digits = 2))
  colnames(df2) <- c("cell_type", "cluster", "percentage")
  print(ggplot(df2, aes(x = cluster, y = percentage))+
    geom_bar(stat = "identity", aes(fill = cell_type))+
    theme_bw()+
    coord_flip()+
    scale_x_discrete(limits = rev(levels(df2$cluster))))
}
```

#### Cell phase

Plots comparing cell type assignments divided by cell phase assignment with the unsupervised clustering

```{r singleR_and_phase_vs_clustering_plots, fig.width=12, fig.height=12, echo=F}
for (g.b in resolutions) {
  res = gsub("harmsnn_res.", "", g.b)
  cat("Resolution : ", res, "\n")
  df <- as.data.frame(round(prop.table(table(so@meta.data[[g.b]], interaction(so$Phase, so$pruned.labels_singleR_fine)), margin = 2)*100, digits = 2))
colnames(df) <- c("cluster", "phase_cell_type", "percentage")
print(ggplot(df, aes(x = phase_cell_type, y = percentage))+
  geom_bar(stat = "identity", aes(fill = cluster))+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(df$phase_cell_type))))
  df2 <- as.data.frame(round(prop.table(table(interaction(so$Phase, so$pruned.labels_singleR_fine), so@meta.data[[g.b]]), margin = 2)*100, digits = 2))
colnames(df2) <- c("phase_cell_type", "cluster", "percentage")
print(ggplot(df2, aes(x = cluster, y = percentage))+
  geom_bar(stat = "identity", aes(fill = phase_cell_type))+
  theme_bw()+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(df2$cluster))))
}
```

## Session info

```{r, echo=F}
sessionInfo()
```
