---
title: "InDen T cells | Post-processing after Harmony integration"
author: "Giorgio Gonnella"
date: 2024-06-09
output:
  html_document:
    theme: sandstone
params:
  pfx: "/srv/baia/"
  libpath: "/home/giorgio/R-pkg-Seurat5/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F, warning=F, echo=F)
.libPaths(c(params$libpath, .libPaths()))

library(Seurat)
library(RColorBrewer)
library(scProportionTest)

prj_root <- paste0(params$pfx, "inden/")
analysis_root <- paste0(prj_root, "analysis-1/")

this_step_dir <- paste0(analysis_root, "steps/004.0.P.T.integration.harmony/")
if (!dir.exists(this_step_dir)) {
  stop("Current step directory does not exist: ", this_step_dir)
}

var_dir <- paste0(this_step_dir, "results/vars/")
if (!dir.exists(var_dir)) {
  stop("Input variable directory does not exist: ", var_dir)
}
```

```{r load, echo=F}
so <- readRDS(paste0(var_dir, "integrated_so.raw.rds"))
```

# Sample IDs

Create simpler sample IDs, saved under the `sample` column
(e.g. `TC_SC03_S1T_so` -> `S1T`). Sample will be a factor.

```{r create_sample_column, echo=T}
so$sample <- sub("TC_SC03_(S\\d+T)_so", "\\1", so$orig.ident)
levels_sample <- sub("TC_SC03_(S\\d+T)_so", "\\1", levels(so$orig.ident))
so$sample <- factor(so$sample, levels=levels_sample)
```

# Sample Name

```{r create_sample_name_column, echo=T}
sample_to_sample_name = c("S1T" = "H1",
                          "S2T" = "H2",
                          "S3T" = "H3",
                          "S5T" = "H4",
                          "S6T" = "H5",
                          "S7T" = "H6",
                          "S8T" = "S1",
                          "S9T" = "S2",
                          "S10T" = "S3",
                          "S11T" = "S4",
                          "S12T" = "S5",
                          "S13T" = "S6")
so$sample_name <- factor(unname(sample_to_sample_name[so$sample]),
                         levels = c("S1", "S2", "S3", "S4", "S5", "S6",
                                    "H1", "H2", "H3", "H4", "H5", "H6"))
```

# Cell types

## Rename unchar celltype label

Rename `unchar` to `Uncharacterized` in the main and fine celltype labels.

```{r rename_unchar_to_uncharacterized, echo=F}
so$pruned.labels_singleR_main <-
  gsub("unchar", "Uncharacterized",
       so$pruned.labels_singleR_main)
so$pruned.labels_singleR_fine <-
  gsub("unchar", "Uncharacterized",
       so$pruned.labels_singleR_fine)
```

## Sum-up gamma-delta T cells

Sum-up `Vd2 gd T cells` and `Non-Vd2 gd T cells` into `Gamma-delta T cells`.

```{r sum_up_gd, echo=TRUE}
so$pruned.labels_singleR_fine <-
  ifelse(so$pruned.labels_singleR_fine == "Vd2 gd T cells",
         "Gamma-delta T cells",
         so$pruned.labels_singleR_fine)
so$pruned.labels_singleR_fine <-
  ifelse(so$pruned.labels_singleR_fine == "Non-Vd2 gd T cells",
         "Gamma-delta T cells",
         so$pruned.labels_singleR_fine)
```

## Group and sort fine celltype labels

Create groups of fine celltype labels (CD4, CD8 and others) and sort them in
a meaningful order

```{r fctype_groups, echo=TRUE}
so@misc$fctype_group <- list()
so@misc$fctype_group[["CD4+"]] <-
  c("Naive CD4 T cells", "Th1 cells", "Th1/Th17 cells", "Th17 cells",
    "Th2 cells", "Follicular helper T cells", "T regulatory cells",
    "Terminal effector CD4 T cells")
so@misc$fctype_group[["CD8+"]] <-
  c("Naive CD8 T cells", "Central memory CD8 T cells",
    "Terminal effector CD8 T cells", "Effector memory CD8 T cells")
so@misc$fctype_group[["Other"]] <-
  c("Gamma-delta T cells", "MAIT cells",
    "Natural killer cells", "Uncharacterized")

so$pruned.labels_singleR_fine <-
  factor(so$pruned.labels_singleR_fine,
         levels = c(so@misc$fctype_group[["CD4+"]],
                    so@misc$fctype_group[["CD8+"]],
                    so@misc$fctype_group$Other))

so$cell_group <- NA
for (group in names(so@misc$fctype_group)) {
  so$cell_group <-
    ifelse(so$pruned.labels_singleR_fine %in%
             so@misc$fctype_group[[group]],
           group, so$cell_group)
}
```

## Abbreviated fine celltype labels

Abbreviate fine celltype labels for plots where there is not enough space to
use the full names. The abbreviated cell types are stored in the
column `pruned.labels_singleR_fine.abbrev` and are a factor.

```{r abbreviate_cell_types, echo=TRUE}
so@misc$pruned.labels_singleR_fine.abbrev <-
  levels(so$pruned.labels_singleR_fine)
names(so@misc$pruned.labels_singleR_fine.abbrev) <-
  levels(so$pruned.labels_singleR_fine)

so@misc$pruned.labels_singleR_fine.abbrev[[
    "Naive CD4 T cells"]] <- "CD4+ Naive"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Th1 cells"]] <- "Th1"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Th1/Th17 cells"]] <- "Th1/Th17"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Th17 cells"]] <- "Th17"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Th2 cells"]] <- "Th2"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Follicular helper T cells"]] <- "Tfh"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "T regulatory cells"]] <- "Tregs"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Terminal effector CD4 T cells"]] <- "CD4+ TEMRA"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Naive CD8 T cells"]] <- "CD8+ Naive"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Central memory CD8 T cells"]] <- "CD8+ CM"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Effector memory CD8 T cells"]] <- "CD8+ EM"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Terminal effector CD8 T cells"]] <- "CD8+ TEMRA"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "MAIT cells"]] <- "MAIT"
so@misc$pruned.labels_singleR_fine.abbrev[[
    "Gamma-delta T cells"]] <- "GD"
```

```{r apply_abbreviated_labels, echo=TRUE}
so$pruned.labels_singleR_fine.abbrev <-
  factor(as.vector(so@misc$pruned.labels_singleR_fine.abbrev[
    so$pruned.labels_singleR_fine]),
    levels = so@misc$pruned.labels_singleR_fine.abbrev[
      levels(so$pruned.labels_singleR_fine)])
```

# Clinical condition

Create a column `clinical_condition` with two levels: `Subclinical` and
`Hospitalized`.

```{r create_clinical_condition_column, echo=TRUE}
conditions <- c("Subclinical", "Hospitalized")
hosp_n = as.character(1:7)
so$clinical_condition <-
  ifelse(so$orig.ident %in% paste0("TC_SC03_S", hosp_n, "T_so"),
         conditions[2], conditions[1])
so$clinical_condition <-
  so$clinical_condition <-
    factor(so$clinical_condition, levels = conditions)
```

```{r create_clinical_condition_abbrev, echo=TRUE}
so$clinical_condition.abbrev <- factor(
  substr(as.character(so$clinical_condition), 1, 1),
  levels = substr(levels(so$clinical_condition), 1, 1)
)
```

# Select clustering resolution

Select the clustering resolution to be used for the downstream analysis.

```{r selected_resolution, echo=TRUE}
so@misc$sel_res <- "1.2"
so@misc$sel_res_col <- paste0("harmsnn_res.", so@misc$sel_res)
```

# Color schemes

Define color schemes for the main cell type labels, fine cell type labels,
clinical conditions, samples, clusters and cell cycle phases, in order to have
consistent colors across all plots.

The colors are saved in some columns of the `so@misc` slot:
- `mctype_cols` for the main cell type labels
- `fctype_cols` for the fine cell type labels
- `cond_cols` for the clinical conditions
- `sample_cols` for the samples
- `sample_name_cols` for the sample names
- `cluster_cols` for the clusters
- `phase_cols` for the cell cycle phases

```{r color_scheme_for_main_labels}
so@misc$mctype_cols <-
  colorRampPalette(brewer.pal(6, "Accent"))(
    length(unique(so$pruned.labels_singleR_main)))
names(so@misc$mctype_cols) <-
  levels(so$pruned.labels_singleR_main)
```

```{r color_scheme_for_fine_labels}
so@misc$fctype_cols <- c(
  "Th17 cells" = "#1f78b4",         # dark blue
  "Th2 cells" = "#a6cee3",          # light blue
  "Th1 cells" = "#000080",          # very dark blue
  "Th1/Th17 cells" = "#00bfff",     # bright blue
  "Naive CD4 T cells" = "#238b45",  # darker green
  "T regulatory cells" = "#aaeeaa", # light green
  "Follicular helper T cells" = "#66c2a4", # blue-green
  "Terminal effector CD4 T cells" = "#9acd32", # yellow green

  "Naive CD8 T cells" = "#e31a1c",          # red
  "Central memory CD8 T cells" = "#fb9a99", # light red
  "Terminal effector CD8 T cells" = "#ffa500", # orange
  "Effector memory CD8 T cells" = "#ff69b4",  # pink

  "Vd2 gd T cells" = "#6a51a3",    # dark purple
  "Non-Vd2 gd T cells" = "#9e9ac8", # light purple
  "Gamma-delta T cells" = "#5a5aa8",  # dark purple

  "MAIT cells" = "#d2b48c",       # light brown
  "Natural killer cells" = "#8c564b", # brown

  "Uncharacterized" = "#bdbdbd"    # gray
)
```

```{r color_scheme_for_conditions}
so@misc$cond_cols <- c(
  "Subclinical" = "darkblue",
  "Hospitalized" = "darkred"
)
```

```{r color_scheme_for_samples}
so@misc$sample_cols <-
  colorRampPalette(brewer.pal(8, name = "Accent"))(
                  length(unique(so$sample)))
# switch colors 1-6 to colors 7-12, to match the changed order of samples
so@misc$sample_cols <- so@misc$sample_cols[c(7:12, 1:6)]
names(so@misc$sample_cols) <- levels(so$sample)

so@misc$sample_name_cols <- so@misc$sample_cols
names(so@misc$sample_name_cols) <- levels(so$sample_name)
```

```{r color_scheme_for_clusters}
g.b = so@misc$sel_res_col
so@misc$cluster_cols <-
  colorRampPalette(brewer.pal(8, name = "Set1"))(
              length(unique(so@meta.data[[g.b]])))
names(so@misc$cluster_cols) <-
  levels(so@meta.data[[g.b]])
```

```{r color_scheme_for_phases}
so@misc$phase_cols <- c("#00AFBB", "#E7B800", "#FC4E07")
names(so@misc$phase_cols) <- c("G1", "G2M", "S")
```

```{r color_scheme_for_plots_background_divided_by_celltype_group}
so@misc$ctg_backcols <-
  c("CD4+" = "#ddffdd", "CD8+" = "#ffcc99", "Other" = "#e6ccff")
```

# Remove outliers

Remove cells in cell types with less than 10 cells and clusters with less than
10 cells.

```{r remove_celltype_outliers, echo=TRUE}
min_cells_per_celltype <- 10
celltype_sizes <- table(so$pruned.labels_singleR_fine)
small_celltypes <- names(celltype_sizes[celltype_sizes < min_cells_per_celltype])
so$in_small_celltype <-
  so$pruned.labels_singleR_fine %in% small_celltypes
so <- subset(so, in_small_celltype == FALSE)
so <- subset(so, pruned.labels_singleR_fine != "Uncharacterized")
so$pruned.labels_singleR_fine <-
  droplevels(so$pruned.labels_singleR_fine)
so$pruned.labels_singleR_fine.abbrev <-
  droplevels(so$pruned.labels_singleR_fine.abbrev)
```

```{r remove_small_clusters, echo=TRUE}
# remove cells in clusters which have less than 10 cells:
min_cluster_size = 10
cluster_sizes <- table(so@meta.data[[so@misc$sel_res_col]])
small_clusters <- as.integer(names(cluster_sizes[cluster_sizes < min_cluster_size]))
so$in_small_cluster <-
  so@meta.data[[so@misc$sel_res_col]] %in% small_clusters
so <- subset(so, in_small_cluster == FALSE)
so$in_small_cluster <- NULL
so@meta.data[[so@misc$sel_res_col]] <-
  droplevels(so@meta.data[[so@misc$sel_res_col]])
```

# Cell proportions permutations test

Perform a permutation test to assess the differences in cell type proportions.

```{r cell_proportions_permutation_test, echo=TRUE}
so@misc$proptest <- permutation_test(sc_utils(so),
                       cluster_identity = "pruned.labels_singleR_fine.abbrev",
                       sample_1 = "Subclinical", sample_2 = "Hospitalized",
                       sample_identity = "clinical_condition",
                       n_permutations = 1000)
```

## Saving data

```{r save, echo=TRUE}
saveRDS(so, paste0(var_dir, "integrated_so.rds"))
```

## Session info

```{r, echo=F}
sessionInfo()
```
